Towards Detecting Stalkerware Through Network Traces

### 2019-03-19 paper/main.tex
Introduction and future work

Although stalkerware is superficially similar to other types of malware, it has 
a few key differences. First, stalkerware is meant to be installed 
deliberately, by someone with physical access to the device that is to be 
monitored. In the context of intimate partner abuse, this presents an enormous 
threat. An abuser will often know or be able to coerce passwords to the device 
from the victim, making traditional defense measures irrelevant. Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Malware is therefore much more likely to 
attempt to obfuscate its network traffic. Third, stalkerware falls for the most 
part into two categories: apps that are explicitly advertised for surveilling 
spouses and partners, and apps that have a different stated use but can be 
repurposed to stalk a target. These are usually marketed for theft prevention, 
parental control of children, and employer control of employees. Previous 
literature refers to these as "dual-use" apps\cite{chatterjee_spyware_2018}. 
Some dual-use apps are ones that the survivor wants to keep installed, such as 
``find my phone'' apps or social media apps that leak the user's location. The 
most troubling example of this is Facebook, which most survivors want to keep 
so that they can contact their support network for help.  
Determining if a particular dual-use app is being used for 
intimate partner surveillance (IPS) presents an enormous challenge. Given the 
wide spectrum of apps that can be used to facilitate IPS and the limited scope 
of this project, we chose to focus on more overt spyware.

Largely due to the differences between stalkerware and other malware, most 
available information on the pervasiveness of stalkerware is anecdotal and 
piecemeal. Small-scale studies on how widespread spyware is do exist, but they 
have limitations. For example, in a survey of 72 
domestic violence shelters performed by NPR in 2014, 85\% said they worked with 
survivors whose abusers had tracked their location using their smartphone's 
GPS. 75\% said they are working with clients who have had their conversations 
monitored by hidden stalkerware apps.\cite{shahani_smartphones_nodate} However, 
this study does not report how many individual clients experienced being 
digitally stalked. Additionally, it and similar work relies on 
reports from domestic violence shelters and health workers, which does not give 
a complete picture of the situation. Little to no data exist regarding spyware 
use on IPA victims who never visit a shelter. Traditionally, antivirus vendors 
collect metrics about 
malicious software, but antivirus often does not detect spyware 
\cite{chatterjee_spyware_2018}. Alarmingly, the stalkerware 
websites themselves claim to have hundreds of thousands of downloads per app, 
but these numbers are not likely to be trustworthy. The goal of this work is 
therefore to identify a means of identifying stalkerware in the wild so that it 
can be enumerated. We collected the network signatures of 24 stalkerware apps 
and showed that they could be easily distinguished from ordinary smartphone 
network traffic. We also propose a simple method of generalizing our signatures 
to spyware that we did not manually examine, and evaluate its robustness.

@@

Wiretapping the phones was accomplished by setting up our personal computers as 
WiFi hotspots. Since we controlled the passwords required to connect to our 
networks, and Windows alerts the user when a device connects to its mobile 
hotspot, we were able to ensure that we did not tap any traffic other than that 
of the test phones.

@@

\subsection{Network Signatures}

@@

>>>>

Stalkerware is not as widely studied as other types of malware, and much less 
information is available regarding it. Although it is superficially 
similar to other types of malware, there exist a few key differences. First, 
stalkerware is meant to be installed 
deliberately, by someone with physical access to the device that is to be 
monitored. In the context of intimate partner abuse, this presents an enormous 
threat. An abuser will often know or be able to coerce passwords to the device 
from the victim, making traditional defense measures irrelevant. This is one 
example of how traditional assumptions about security often do not apply in the 
context of IPV (intimate partner violence). Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Both malware and stalkerware need to hide 
their presence on the target device, but malware is much more likely to 
attempt to obfuscate its network traffic. Stalkerware does not: it does not 
need to. Most antivirus does not detect apps capable of monitoring a target as 
potentially dangerous \cite{chatterjee_spyware_2018}. This presents an 
opportunity to glean information about stalkerware that is not available in the 
study of other forms of malware. Third, stalkerware falls 
for the most part into two categories: apps that are explicitly advertised for 
surveilling spouses and partners, and apps that have a different stated use but 
can be repurposed to stalk a target. These are usually marketed for theft 
prevention, finding lost devices, parental control of children, and employer 
control of employees. Previous literature refers to these as "dual-use" 
apps\cite{chatterjee_spyware_2018}. Some dual-use apps are ones that the 
victim wants to keep installed, such as 
``find my phone'' apps or social media apps that leak the user's location. The 
most troubling example of this is Facebook, which most victims want to keep 
so that they can contact their support network for help.  Determining if a 
particular dual-use app is being used for intimate partner surveillance (IPS) 
presents an enormous challenge. Given the wide spectrum of apps that can be 
used to facilitate IPS and the limited scope of this project, we chose to focus 
on more overt spyware.

Largely due to the differences between stalkerware and other malware, most 
available information on the \textit{pervasiveness} of stalkerware is anecdotal 
and piecemeal. Small-scale studies on how widespread spyware is do exist, but 
they have limitations. For example, in a survey of 72 
domestic violence shelters performed by NPR in 2014, 85\% said they worked with 
survivors whose abusers had tracked their location using their smartphone's 
GPS. 75\% said they have worked with clients who have had their conversations 
monitored by hidden stalkerware apps.\cite{shahani_smartphones_nodate} However, 
this study does not report how many individual clients experienced being 
digitally stalked. Additionally, it and similar work relies on 
reports from domestic violence shelters and health workers, which do not give 
a complete picture of the situation. Little to no data exist regarding spyware 
use on IPV victims who never visit a shelter. Traditionally, antivirus vendors 
collect metrics about 
malicious software, but antivirus often does not detect spyware 
\cite{chatterjee_spyware_2018}. Alarmingly, the stalkerware 
websites themselves claim to have hundreds of thousands of downloads per app, 
but these numbers are not likely to be trustworthy. The goal of this work is 
therefore to identify a means of identifying stalkerware in the wild so that it 
can be enumerated. We identified unique signatures in the traffic of 24 
stalkerware apps, by examining a traffic features such as domain names, HTTP 
header fields, protocol, and server name indicators (SNIs). For a complete 
list, see section \ref{network_signatures}. These signatures can be used by 
antivirus software and IDSs to determine how widespread stalkerware is. We show 
that these signatures are distinguishable from ordinary smartphone 
network traffic. We also propose a simple method of generalizing our signatures 
to spyware that we did not manually examine, and evaluate its robustness.

@@

Wiretapping the phones was accomplished by setting up our personal computers as 
WiFi hotspots. Since we controlled the passwords required to connect to our 
networks, and Windows alerts the user when a device connects to its mobile 
hotspot, we were able to ensure that we did not tap any traffic other than that 
of the test phones. 
\section{Results}

\subsection{Generalizing Signatures}

We picked our three keywords because of xyz reasons outlined in the checkpoint

In the subset of apps we studied, the words “spy,” “find,” and “track” appeared 
frequently in various fields of the traffic. The most common place to find them 
was in the destination domain, which we found using Wireshark’s reverse DNS 
lookup tool. 12 of the 26 apps that we eventually found signatures for 
contained these spyware keywords in their domains. 
For the apps whose traffic doesn’t contain “spy,” “find,” or “track,” we will 
next check for the app name and manufacturer name or substrings of these two. 
Using these indicators has a significant disadvantage - we must be certain that 
the app is stalkerware and that the manufacturer doesn’t make anything except 
stalkerware. In the case of smaller apps and reskinned apps, this may actually 
be a reasonable assumption - manufacturer names are generated to fool the 
Google Play store into thinking the app is new, not to identify a real company, 
and are therefore unique by necessity. We note that the two apps we identify 
only by manufacturer are made by GPSWOX and GeoZilla. GPSWOX makes 
location-tracking apps of all kinds, some of which may be difficult to use as 
stalkerware, but GeoZilla’s only apps are all designed to track a phone. Since 
only two apps were identified this way, the false positive rate should be 
investigated further.
We identified the following generalized signatures that can be used to identify 
most of our apps given nothing but information available on the Google Play 
store. The second step of our analysis will be to search for these general 
signatures in our personal phones’ traffic captures. This will give a better 
idea of whether it is possible to use general signatures without generating a 
prohibitive number of false positive stalkerware detections.

@@

\subsection{Network Signatures}
\label{network_signatures}

@@

Previously, we had expressed an interest in seeing if the signatures we found 
by hand could identify more apps without manual inspection. There are (as far 
as we can see) two possible methods of doing this. The first is to 
auto-generate signatures based on a list of keywords that commonly appear in 
domains, SNIs, User-Agents, and HTTP host names for stalkerware traffic. These 
keywords might include the names of the apps, substrings that appear in the app 
names, and words that seem relevant to tracking apps. A more complicated method 
would be to set up bots that can download potential spyware by searching 
various terms on the app store like “tracker app.” The bots would have to run 
the apps and monitor their traffic, identifying app traffic by eliminating 
traffic from other sources. The remaining traffic would belong to the app, and 
the domains, SNIs, user-agents, and any other available identifying features 
could be recorded. However, the most difficult problem in this case becomes 
determining which apps are actually stalkerware on the app store, which is 
beyond the scope of this project.
